---
output: github_document
title: "Process Accelerometer Data from NHANES 2003-2006"
author: "Dane Van Domelen <br> vandomed@gmail.com"
date: "`r Sys.Date()`"
bibliography: bibliography.bib
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

[![Build Status](https://travis-ci.org/vandomed/nhanesaccel.svg?branch=master)](https://travis-ci.org/vandomed/nhanesaccel)

```{r setup, include = FALSE}

knitr::opts_chunk$set(
  collapse = TRUE,
  eval = TRUE,
  comment = "#>",
  fig.path = "README-",
  message = FALSE
)

# Load packages
library("dvmisc")
#library("printr")
library("knitr")
library("nhanesaccel")
```


## Introduction

The National Health and Nutrition Examination Survey (NHANES) included physical activity monitoring to their 2003-2004 and 2005-2006 study cycles. Altogether, almost 15,000 participants age $\ge$ 6 years wore an ActiGraph accelerometer at the hip for 7 consecutive days. The data are freely available to download from the [NHANES website](https://wwwn.cdc.gov/nchs/nhanes/continuousnhanes/default.aspx?BeginYear=2003). You don't have to submit a proposal, you don't have to find a study investigator to ~~leach authorship~~ help you understand the data, none of that. I think the CDC deserves a lot of credit for this setup.

The raw accelerometer data needs to be processed to generate meaningful physical activity varaibles, and indeed a SAS script was provided by the National Cancer Institute for this purpose [@ncisas]. But it only calculates a few variables, it isn't easily modifiable, and it takes a long time to run. So I developed **nhanesaccel** to process the data in R much faster and with more options.


## Install

The package includes the raw data and functions for processing it. It's way too big for CRAN because of the raw data, so it's only on GitHub. To install it, you can run:

```{r eval = FALSE}
library("devtools")
install_github("vandomed/nhanesaccel")
```

If this doesn't work, you might try installing **accelerometry** [@accelerometry] from CRAN first with `install.packages("accelerometry")`. If you still have problems, e-mail me at <vandomed@gmail.com> and I can try to help.


## Usage

#### Defaults

The primary function in **nhanesaccel** is *process_nhanes*. If you call it without specifying anything, it will process the 2003-2006 data with default data-processing parameters (i.e. my personal preferences).

```{r eval = TRUE}
df <- process_nhanes()
```

As you can see from the messages, it processed data from both 2003-2004 and 2005-2006. To see what variables it generated:

```{r}
names(df)
```

I won't go through all of them, but `seqn` is the subject ID number, `valid_days` is the number of days out of 7 that were deemed suitable for analysis (the subject wore the accelerometer long enough), `valid_min` is the average number of wear time minutes per day, and `counts` and `cpm` are two common indicators of physical activity volume. The `wtmec2yr_adj` and `wtmec4yr_adj` variables are the adjusted MEC weights that you'll want to use for statistical analysis.

Let's look at the relationship between `valid_min` and `counts` in the full dataset.

```{r}
library("ggplot2")
p <- ggplot(df, aes(x = valid_min, y = counts)) + 
  geom_point() + 
  xlim(0, 1440) + ylim(0, 1750000) + 
  geom_smooth(method = "lm") + 
  labs(title = "Counts vs. Wear Time", x = "Wear time (min/day)", y = "Counts/day") + 
  theme_bw(base_size = 14)
p
```

Notice that `valid_min` ranges from 600 to 1440. The lower bound is due to the default requirement that only days with 600 minutes (10 hours) of wear time be included, and the upper bound, 1440, is the number of minutes in 24 hours. The fact that some subjects average 1440 minutes of wear time means either they didn't sleep (not likely!) or maybe forgot to remove the device before bed and had small movements while sleeping.

It looks like counts ranged from about 0 all the way up to around 1.5 million, which is very high but plausible considering that roughly 10,000 counts per minute can be achieved during intense physical activity.

As you would expect, participants who wore the device longer tended to accumulate more counts. That's why we typically adjust for wear time when fitting regression models with accelerometer-derived variables.

#### Mimicking the NCI's SAS programs

I figured a lot of users would want to replicate the NCI's SAS programs, so I wrote *process_nhanes* to allow for that. Rather than specifying about 30 separate inputs, you can simply use the `nci_methods` input as a shortcut. To illustrate with just the 2003-2004 data:

```{r}
df.0304 <- process_nhanes(waves = 1, nci_methods = TRUE)
```

Obviously I don't fully agree with the NCI's methods (otherwise they would be the defaults!), but a lot of it is subjective. One thing I don't like is that they use a really liberal algorithm for non-wear detection, which I suspect tends to misclassify sedentary time as non-wear. 

#### More variables, different settings

You can read about all of the different options in the help file (run `?process_nhanes`), but I'll give one example of adjusting the default parameters. To process the 2003-2004 data with a 45-minute non-wear algorithm, requiring at least 4 valid days of data, and requesting a bunch of variables, you could run:

```{r}
df2 <- process_nhanes(waves = 1, nonwear_window = 45, valid_days = 4, brevity = 3)
```

#### Per-person vs. per-day data

The default behavior of *process_nhanes* is to return a per-person file with physical activity variables averaged across all valid days for each subject. If you want a per-day file instead, in long format, you can set `return_form = "daily"` rather than the default `return_form = "averages"`. For a list containing both datasets, you can set `return_form = "both"`.


## Merging other datasets and fitting models

This is a little beyond the scope of the README, but I would suggest my **nhanesdata** package [@nhanesdata] for accessing the other NHANES datasets and **survey** [@survey1; @survey2] for fitting models while accounting for the survey design of NHANES. You can install **nhanesdata** with `install_github("vandomed/nhanesdata")` and **survey** with `install.packages("survey")`.

Here is a simple example of merging in two datasets (demographics and body measures) to our first `df` dataset and fitting a model for waist circumference vs. sex and average daily CPM in adults age 18-29.

```{r}
# Load packages
library("nhanesdata")
library("survey")
library("dplyr")

# Merge in demographics and body measures datasets and recode sex
names(df) <- toupper(names(df))
df <- df %>% 
  dplyr::left_join(full_join(demo_c, demo_d)) %>%
  dplyr::left_join(full_join(bmx_c, bmx_d)) %>%
  dplyr::mutate(sex = factor(RIAGENDR, levels = c(2, 1), labels = c("Female", "Male")))

# Specify survey design object
design <- svydesign(
  data = subset(df, INCLUDE == 1),
  ids = ~SDMVPSU, 
  weights = ~WTMEC4YR_ADJ, 
  strata = ~SDMVSTRA, 
  nest = TRUE
)

# Fit waist circumference vs. sex and average CPM
fit <- svyglm(BMXWAIST ~ sex + CPM, 
              design = subset(design, RIDAGEYR %in% 18: 29 & ! is.na(BMXWAIST)))
summary(fit)$coef
```

We see that males average  4.44 cm larger waists than females, and CPM (our measure of physical activity volume) is inversely associated with waist circumference, as expected. Note that the slopes for males and females were not statistically different (interaction P = 0.67).

Here is a scatterplot to go along with the fitted model:

```{r}
p <- ggplot(subset(df, RIDAGEYR %in% 18: 29), 
            aes(x = CPM, y = BMXWAIST, color = sex)) + 
  geom_point() + guides(color = guide_legend(title = "Sex")) + 
  ylim(0, 175) + 
  geom_abline(intercept = fit$coef[1], slope = fit$coef[3], color = "#F8766D") +
  geom_abline(intercept = fit$coef[1] + fit$coef[2], slope = fit$coef[3], color = "#00BFC4") + 
  labs(title = "Waist Circumference vs. Average Daily CPM", 
       x = "CPM", y = "Waist circumference (cm)") + 
  theme_bw(base_size = 14)
p
```


## Shiny app

Part of my motivation for developing **nhanesaccel** was to allow researchers without a SAS license (or knowledge of SAS) to process the data. But what about people who don't use R either?

I decided to convert *process_nhanes* into a Shiny app, so that users could just visit a website, select various options within a GUI, and ultimately process the data and export an analysis file.

I've written the app, but it isn't online yet. For now, you have to launch it from R by running:

```{r eval = FALSE}
process_nhanes_app()
```


## References
